---
title: "Example HMP analysis"
author: "Aram Avila-Herrera"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(fig.width = 8, fig.height = 8,
                      collapse = TRUE, comment = '#>'
                      )

```

The `novelgeneshmp` package accompanies supplemental methods for
**NOVELGENES_TITLE_HERE** by **AUTHORLIST**. It contains functions to
facilitate tests for association between subject metadata from the Human
Microbiome Project (HMP) and under-annotated protein families investigated in
the manuscript (FUnkSFAMs).

## Finding associations between FUnkSFAM abundance and subject metadata in Human Microbiome Project samples


```{r, warning = FALSE, messages = FALSE}

library(novelgeneshmp)

```

I like using `dplyr` and `magrittr` for manipulating data.frames, and
`ggplot2` for plotting.

```{r, warning = FALSE, messages = FALSE}

library(dplyr)
library(magrittr)
library(ggplot2)

```

### Configuration file

File paths and other options are specified in a `json` formatted configuration file
that drives the majority of data 'clean-up' and output.

```{r}

# Specify configuration filename
config_fn = system.file("extdata",
                        "novelgenes_configv.json",
                        package = "novelgeneshmp"
                        )

# Uses jsonlite::fromJSON to load the config into a list
config_l = load_config(config_fn)

```

#### Example config:

```json
{
    "DCM_fn"             : "/path/to/phs000228.v3.pht001187.v3.p1.c1.EMMES_HMP_DCM.HMP.txt",
    "DSU_fn"             : "/path/to/phs000228.v3.pht002156.v1.p1.c1.EMMES_HMP_DSU.HMP.csv",
    "DEM_fn"             : "/path/to/phs000228.v3.pht002158.v1.p1.c1.EMMES_HMP_DEM_ENR.HMP.txt",
    "DTPDHXDVD_fn"       : "/path/to/phs000228.v3.pht002157.v1.p1.c1.EMMES_HMP_DTP_DHX_DVD.HMP.csv",
    "GTV_fn"             : "/path/to/phs000228.v3.pht001193.v3.p1.c1.EMMES_HMP_GTV.HMP.txt",

    "counts_fn"          : "/path/to/HMP_funksfams_counts.csv",

    "project_catalog_fn" : "/path/to/HMP_project_catalog.csv",
    "srs2randsid_fn"     : "/path/to/SRSID2RANDSID.tsv",
    "srs2visno_fn"       : "/path/to/ppAll_V35_map.txt",

    "count_threshold"    : 1,

    "span67_fn"          : "/path/to/span67.tsv",

    "output_dir"         : "./output_dir/",
    "inspect_dir"        : "./inspect_dir/"
}

```

#### Subject metadata

Specify where the subject metadata is. This data is protected and may be requested from
[dbGaP](http://www.ncbi.nlm.nih.gov/projects/gap/cgi-bin/study.cgi?study_id=phs000228.v3.p1).
We will not redistribute it.


| config attribute | data          | variables                                          |
|:-----------------|:--------------|:---------------------------------------------------|
| `DCM_fn`         | DCM           | Medications                                        |
| `DSU_fn`         | DSU           | Diet                                               |
| `DEM_fn`         | DEM           | Birthcountry                                       |
| `DTPDHXDVD_fn`   | DTP DHX DVD   | BMI, smoker status, occupation, etc...             |
| `GTV_fn`         | GTV           | Used for `RANDSID` to (`SRSID`, `VISNO`) mapping   |

#### FUnkSFAM read counts (abundance -> presence-absence)

`counts_fn`:

- CSV formatted
- Each row is a FUnkSFAM
- First column is a FUnkSFAM
- Other columns are counts for each `SRSID`

`count_threshold`:

- FUnkSFAMs are considered present if there are more than `count_threshold` read
  counts

#### Auxiliary data and data mappings

| config attribute         | data                                       | variables                                          |
|:-------------------------|:-------------------------------------------|:---------------------------------------------------|
| `project_catalog_fn`     | HMP project catalog                        | Body site and subsites                             |
| `srs2randsid_fn`         | A mapping from SRSID to RANDSID            | `SRSID`, `RANDSID`                                 |
| `srs2visno_fn`           | A mapping from SRSID to VISNO              | `SRSID`, `VISNO` *only contains VISNO = {1, 2, 3}* |
| `span67_fn`              | Number of annotations across databases     | `FUNKID`, `N_annotations`                          |

#### Output

These attributes aren't required, but they are handy.

- `output_dir`: Put results here
- `inspect_dir`: Put intermediate files here

In these examples, the directory names are expected to end in a `/`.
Directories can be created if they don't exist with `dir.create(path,
showWarnings = FALSE)`.

### Data preprocessing.

Subject metadata is joined and formatted into two tables. One that contains
subject metadata that was collected at each visit, and one that contains
subject metadata that is not expected to change across visit (e.g. birth
country).

```r

phenos_l = load_and_clean_phenotypes(config_l)

# Loads counts and convert to presence-abscence
pres_df = prepare_abundance(config_l)
srs_map = prepare_map(config_l)  # maps `SRSID` to (`RANDSID`, `VISNO`)

# get `VISNO`s that are mappable to samples
mappable_visnos = get_mappable_visnos(srs_map)  # Only primary visits in this case

# Inspect printed output for DCMCODE_XX, etc...
check_mappable_phenos(phenos_l, mappable_visnos)

# make phenotype tables, using `check_mappable_phenos` output.
ph_vm = get_visno_mappable_pheno_df(phenos_l, mappable_visnos)     # complicated functions
ph_vu = get_visno_unmappable_pheno_df(phenos_l, mappable_visnos)   # with hardcoded phenotypes


# merge phenotype tables with presence-abscence tables
visno_data_df = Reduce(inner_join, list(pres_df, srs_map, ph_vm))
novisno_data_df = Reduce(inner_join, list(pres_df, srs_map, ph_vu))

```

Because the data is protected, our example proceeds with randomly generated
placeholder data.


```{r}

# Specify where cleaned data would be
visno_data_fn = system.file("extdata",
                            paste0(config_l$inspect_dir, "visno_mapped_samples_cleaned.tsv"),
                            package = "novelgeneshmp"
                            )
novisno_data_fn = system.file("extdata",
                            paste0(config_l$inspect_dir, "visno_unmapped_samples_cleaned.tsv"),
                            package = "novelgeneshmp"
                            )

# Load the cleaned data
visno_data_df = load_tsv(visno_data_fn)
novisno_data_df = load_tsv(novisno_data_fn)

```

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = 'hide'}

config_l$span67_fn = system.file("extdata",
                                 "span67.tsv",
                                 package = "novelgeneshmp"
                                 )

```

### Variable summaries

Get summaries for phenotypes (subject metadata) and FUnkSFAM abundance
(presence-absence).

```{r}

# Number of samples per body (sub)site for data that maps to VISNO of origin
# and for data that is not expected to change across VISNOs
nsamples_by_bodysite_df = bind_rows(   # dplyr::bind_rows() is like rbind()
    calc_nsamples_per_group(visno_data_df, 'HMP_BodySite') %>%
        mutate(VISNO_MAPPED = TRUE),
    calc_nsamples_per_group(novisno_data_df, 'HMP_BodySite') %>%
        mutate(VISNO_MAPPED = FALSE)
    )

# FUnkSFAM variation statistics by body (sub)site.
ff_stats_bss_df = calc_FFvariation_per_group(novisno_data_df, 'HMP_BodySubsite')

# Phenotype (subject metadata variable) variation statistics.
ph_stats_bs_df = bind_rows(calc_PHvariation_per_group(visno_data_df, 'HMP_BodySite'),
                           calc_PHvariation_per_group(novisno_data_df, 'HMP_BodySite')
                           )

```

We can view the summaries:

```{r}

head(nsamples_by_bodysite_df)
head(ph_stats_bs_df)
head(ff_stats_bss_df)

```

and/or save them to file, perhaps somewhere in `inspect_dir`:

```r

save_tsv(nsamples_by_bodysite_df,
         paste0(config_l$inspect_dir, 'nsamples_by_bodysite.tsv')  # put it in `inspect_dir`
         )

```

or plot them:

```{r, warning = FALSE}

plot_FFentropy_by_group(ff_stats_bss_df, 'HMP_BodySubsite')

```

## Fit logistic regression models and test for coefficients significantly larger than zero

Coefficients from a logistic regression were used to identify likely
associations between each subject variable and presence of each FUnkSFAM. The
models account for geographic location (`SITE` variable in HMP) and were fit for
each body site (or separately sub-site) and for each pair of subject variable
and FUnkSFAM.

To cut down on computate time, `do_glm_tests()` calls
`prefilter_by_arbitrary_statistic()` to skip model fitting if there is not enough
variation in the subject variable (e.g. everyone is a vegetarian).

```{r, warning = FALSE}

res_by_bodysite_df = bind_rows(
    do_glm_tests(visno_data_df, 'HMP_BodySite'),
    do_glm_tests(novisno_data_df, 'HMP_BodySite')
    )
res_by_bodysubsite_df = bind_rows(
    do_glm_tests(visno_data_df, 'HMP_BodySubsite'),
    do_glm_tests(novisno_data_df, 'HMP_BodySubsite')
    )

head(res_by_bodysite_df)
head(res_by_bodysubsite_df)

```

## Post-filtering and p-value correction

To avoid a high penalty for multiple testing, we *a priori* decide to focus on
a specific subset of tests.

### Filtering functions

| name                                             | removes tests for...                                             |
|:-------------------------------------------------|:-----------------------------------------------------------------|
| `filter_results_by_FFentropy_per_group`          | mostly present or absent FUnkSFAMs                               |
| `filter_results_by_funksfam_annotation`          | phylogenetically narrow FUnkSFAMs or with 10 or more annotations |
| `filter_results_by_per_test_arbitrary_statistic` | subject variable with few differeing values                      |
| `filter_results_by_interesting_DCMCODE`          | DCMCODEs we decided not to investigate                           |
| `filter_unneeded_tests`                          | `SITE` and `(Intercept)` coefficients                            |
| `filter_results_all`                             | all of the above                                                 |


#### FUnkSFAM filtering

##### `filter_results_by_funksfam_annotation`

FUnkSFAMs of span 6 and 7 were selected for analysis if they had less than 10
annotations across protein family data bases. Of these, FUnkSFAMs were included
only if their presence varied appreciably among samples within body sites (or
sub-sites) in order to exclude FUnkSFAMs that were present or absent from a
majority of samples. See `span67_fn`.

##### `filter_results_by_FFentropy_per_group`

We used entropy (`entropy` package in CRAN) to quanitfy our uncertainty about
each FUnkSFAM's presence across samples. FUnkSFAMs in the top quartile by entropy
were considered for analysis.

#### Metadata filtering

##### `filter_results_by_per_test_arbitrary_statistic`

Subject variables were excluded if they did not vary
appreciably among samples within body sites (or sub-sites). Subject variables
were required to have at least two values with more than four observations. For
example, we would choose to keep `DSUDIET` for analysis if:

> There are 5 samples with `DSUDIET = 1`, 6 with `DSUDIET = 2`, 3 with `DSUDIET = 3`

but exclude from analysis if

> There are 5 samples with `DSUDIET = 1`, 4 with `DSUDIET = 2`, 3 with `DSUDIET = 3`

### Adjusting p-values for false discovery

```{r}

# apply all filters
res_bss_fil_df = filter_results_all(res_by_bodysubsite_df, ff_stats_bss_df,
                                    'HMP_BodySubsite', config_l
                                    )

# calls p.adjust(p, method = 'fdr') and sorts the results table
res_bss_df = res_bss_fil_df %>% adjust_pvalues() %>% format_final_results()

head(res_bss_df)

```

```r

# Create output directory, and save results
dir.create(config_l$output_dir, showWarnings = FALSE)
save_tsv(res_bss_df, paste0(config_l$output_dir, 'adjusted_results_by_bodysubsite.tsv'))

```
