---
title: "Example Analysis"
author: "Aram Avila-Herrera"
date: "`r Sys.Date()`"
output:
    html_document:
        toc: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(fig.width = 8, fig.height = 8)

```

<style type="text/css">
table {

    max-width: 80%;

}
</style>


This is a short vignette that demonstrates the analysis workflow that
accompanies the manuscript **(Add doi here?)**.

## Finding associations between subject metadata and FUnkSFAM abundance

The `novelgenes` package contains functions to facilitate tests for association
between subject metadata from the Human Microbiome Project (HMP) and
under-annotated protein families.

```{r, warning = FALSE, messages = FALSE}

library(devtools)
devtools::load_all()
# library(novelgenes)  # soon

```

NB. I like using `dplyr` and `magrittr` for manipulating data.frames, and
`ggplot2` for plotting.

```{r, warning = FALSE, messages = FALSE}
library(dplyr)
library(magrittr)
library(ggplot2)
```

### Config --- (`json` formatted)

A `json` formatted configuration file drives most of the data cleanup,
including some filtering steps and specifying output directories.

```{r}

config_fn = system.file("extdata",
                        "novelgenes_configv.json",
                        package = "novelgenes"
                        )
config_l = load_config(config_fn)

```

#### Example config:

```json
{
    "DCM_fn"             : "/path/to/phs000228.v3.pht001187.v3.p1.c1.EMMES_HMP_DCM.HMP.txt",
    "DSU_fn"             : "/path/to/phs000228.v3.pht002156.v1.p1.c1.EMMES_HMP_DSU.HMP.csv",
    "DEM_fn"             : "/path/to/phs000228.v3.pht002158.v1.p1.c1.EMMES_HMP_DEM_ENR.HMP.txt",
    "DTPDHXDVD_fn"       : "/path/to/phs000228.v3.pht002157.v1.p1.c1.EMMES_HMP_DTP_DHX_DVD.HMP.csv",
    "GTV_fn"             : "/path/to/phs000228.v3.pht001193.v3.p1.c1.EMMES_HMP_GTV.HMP.txt",

    "counts_fn"          : "/path/to/HMP_funksfams_counts.csv",

    "project_catalog_fn" : "/path/to/HMP_project_catalog.csv",
    "srs2randsid_fn"     : "/path/to/SRSID2RANDSID.tsv",
    "srs2visno_fn"       : "/path/to/ppAll_V35_map.txt",

    "count_threshold"    : 1,

    "span67_fn"          : "/path/to/span67.tsv",

    "output_dir"         : "./output_dir/",
    "inspect_dir"        : "./inspect_dir/"
}

```

#### Subject variables

| config attribute | data          | variables                                          |
|:-----------------|:--------------|:---------------------------------------------------|
| `DCM_fn`         | DCM           | Medications                                        |
| `DSU_fn`         | DSU           | Diet                                               |
| `DEM_fn`         | DEM           | Birthcountry                                       |
| `DTPDHXDVD_fn`   | DTP DHX DVD   | BMI, smoker status, occupation, etc...             |
| `GTV_fn`         | GTV           | Used for `RANDSID` to (`SRSID`, `VISNO`) mapping   |

#### FUnkSFAM read counts (abundance -> presence-absence)

`counts_fn`:

- Csv formatted
- Each row is a FUnkSFAM
- First column is a FUnkSFAM
- Other columns are counts for each `SRSID`

`count_threshold`:

- FUnkSFAMs considered present if there are more than `count_threshold` read
  counts

#### Other data and mappings

| config attribute         | data                                       | variables                                          |
|:-------------------------|:-------------------------------------------|:---------------------------------------------------|
| `project_catalog_fn`     | HMP project catalog                        | Body site and subsites                             |
| `srs2randsid_fn`         | A mapping from SRSID to RANDSID            | `SRSID`, `RANDSID`                                 |
| `srs2visno_fn`           | A mapping from SRSID to VISNO              | `SRSID`, `VISNO` *only contains VISNO = {1, 2, 3}* |
| `span67_fn`              | Number of annotations across databases     | `FUNKID`, `N_annotations`                          |

#### Output

These attributes aren't required, but they are handy. They can be used to create
directories if they don't exist.

- `output_dir`: Put results here
- `inspect_dir`: Put intermediate files here

### Pre-processing

Coming soon. Generating the raw fake data sets is complicated. In the meantime,
please enjoy a random cleaned example.

### Pre-processed data

Two tables of random fake data formatted to be ready to use are included. Much
of the HMP medical data can only be obtained through a data usage agreement.

After loading the raw data, cleaned tab-separated-value files are typically
placed in `inspect_dir` with other intermediate files and graphs. We can load
these files and start the example from here.

```{r}

visno_data_fn = system.file("extdata",
                            paste0(config_l$inspect_dir, "visno_mapped_samples_cleaned.tsv"),
                            package = "novelgenes"
                            )
novisno_data_fn = system.file("extdata",
                            paste0(config_l$inspect_dir, "visno_unmapped_samples_cleaned.tsv"),
                            package = "novelgenes"
                            )

# Load (random cleaned data)
visno_data_df = load_tsv(visno_data_fn)
novisno_data_df = load_tsv(novisno_data_fn)

```

`span67_fn` is also included.

```{r, warning = FALSE, results = 'hide'}

config_l$span67_fn = system.file("extdata",
                                 "span67.tsv",
                                 package = "novelgenes"
                                 )

```

### Variable summaries

Get summaries for phenotypes (subject metadata) and FUnkSFAM abundance
(presence-absence).

```{r}

# Number of samples per body (sub)site for data that maps to VISNO of origin
# and for data that is not expected to change across VISNOs
nsamples_by_bodysite_df = bind_rows(
    calc_nsamples_per_group(visno_data_df, 'HMP_BodySite') %>%
        mutate(VISNO_MAPPED = TRUE),
    calc_nsamples_per_group(novisno_data_df, 'HMP_BodySite') %>%
        mutate(VISNO_MAPPED = FALSE)
    )

# FUnkSFAM variation statistics by body (sub)site.
ff_stats_bss_df = calc_FFvariation_per_group(novisno_data_df, 'HMP_BodySubsite')

# Phenotype (subject metadata variable) variation statistics.
ph_stats_bs_df = bind_rows(calc_PHvariation_per_group(visno_data_df, 'HMP_BodySite'),
                           calc_PHvariation_per_group(novisno_data_df, 'HMP_BodySite')
                           )

```

We can view the summaries:

```{r}

head(nsamples_by_bodysite_df)
head(ph_stats_bs_df)

```

and/or save them to file, perhaps somewhere in `inspect_dir`:

```r
save_tsv(nsamples_by_bodysite_df,
         paste0(config_l$inspect_dir, 'nsamples_by_bodysite.tsv')
         )

```

or plot them:

```{r, warning = FALSE}

plot_FFentropy_by_group(ff_stats_bss_df, 'HMP_BodySubsite')

```

## Fit models and test for significantly large coefficients

Model each FUnkSFAM's abundance as a function of each subject variable,
accounting for `SITE` and stratified by body site or subsite.

To cut down on computate time, `do_glm_tests()` calls
`prefilter_by_arbitrary_statistic()` to skip modelling if there is not enough
variation in the subject variable (e.g.  everyone is a vegetarian).

```{r, warning = FALSE}

res_by_bodysite_df = bind_rows(
    do_glm_tests(visno_data_df, 'HMP_BodySite'),
    do_glm_tests(novisno_data_df, 'HMP_BodySite')
    )
res_by_bodysubsite_df = bind_rows(
    do_glm_tests(visno_data_df, 'HMP_BodySubsite'),
    do_glm_tests(novisno_data_df, 'HMP_BodySubsite')
    )

head(res_by_bodysite_df)
head(res_by_bodysubsite_df)

```

## Post-filtering and p-value correction

To avoid a high penalty for multiple testing, we *a priori* decide to focus on
a specific subset of tests.

### Filtering functions

| name                                             | removes tests for...                                             |
|:-------------------------------------------------|:-----------------------------------------------------------------|
| `filter_results_by_FFentropy_per_group`          | mostly present or absent FUnkSFAMs                               |
| `filter_results_by_funksfam_annotation`          | phylogenetically narrow FUnkSFAMs or with 10 or more annotations |
| `filter_results_by_per_test_arbitrary_statistic` | subject variable with few differeing values                      |
| `filter_results_by_interesting_DCMCODE`          | keeps a hardcoded set of DCMCODEs                                |
| `filter_unneeded_tests`                          | `SITE` and `(Intercept)` coefficients                            |
| `filter_results_all`                             | all of the above                                                 |

```{r}

res_bss_fil_df = filter_results_all(res_by_bodysubsite_df, ff_stats_bss_df,
                                    'HMP_BodySubsite', config_l
                                    )

res_bss_fil_df %>% adjust_pvalues() %>% format_final_results() -> res_bss_df

#dir.create(config_l$output_dir, showWarnings = FALSE)
#save_tsv(res_bss_df, paste0(config_l$output_dir, 'adjusted_results_by_bodysubsite.tsv'))
head(res_bss_df)

```
